A data-driven model of nucleosynthesis with chemical tagging in latent space

### 2019-05-08 article/mcfa.tex
lotsa wordso

\shorttitle{Chemical tagging in lower-dimensional latent space}
\shortauthors{Casey et al.}

@@

\title{Chemical tagging in lower-dimensional latent space}

@@

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated chemical abundances as being 
statistically independent, despite the nucleosynthesis literature showing that
most elements can be produced by more than one nucleosynthetic pathway.
In this work we introduce a data-driven model for chemical abundances
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and the clustering (e.g., chemical tagging) 
is represented by a mixture model in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the relative weighting (e.g., partial assignment) of each star
to each cluster, and to infer the latent factors that are common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We perform experiments on literature chemical abundances
of metal-poor stars, and those from the second \Galah\ data release. 
We find that \todo{five} latent factors are preferred to explain the detailed 
abundances of metal-poor stars, and \todo{eight} are preferred given \todo{X} abundances
in \Galah\ data. In both experiments we recover latent factors that 
are representative of mean yields expected from the rapid- and slow-neutron
capture process, as well as nucleosynthesis in Type II supernovae. 
We find \todo{two} clusters are preferred in the \Galah\ data of \todo{X} abundances
and \todo{Y} stars, which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. We expect more components
would be preferred given complete abundance inventory of a larger sample of stars.
We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
enabling model parameter estimation in minutes given $\sim$$10^7$
chemical abundance measurements.
\end{abstract}

@@

Here we present a new approach to chemical tagging that allows us to identify 
the latent (unobserved) factors that contribute to the chemical abundances of 
all stars (e.g., nucleosynthetic yields) while simultaneously performing 
clustering in the latent space. Notwithstanding caveats that we will
discuss in detail, this allows us to infer nucleosynthetic yields rather than
strictly prescribe them from models. Moreover, the scale of the clustering
problem reduces by a significant fraction because the clustering is performed in
a lower dimensional latent space instead of the higher dimensional data space.
In Section~\ref{sec:methods} we describe the model and the methods we use to
estimate the model parameters. Section~\ref{sec:experiments} describe the 
experiments performed using generated and real data sets. We discuss the results
of these experiments in Section~\ref{sec:discussion}, including the caveats with
the model as described. We conclude in Section~\ref{sec:conclusion}.

@@

At the maximization step we update our estimates of the parameters,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
The updated parameters estimates are found by setting the second derivative
of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
the parameter values. In doing so this guarantees that every updated
estimate of the model parameters is guaranteed to increase the log likelihood.
Although there are no guarantees against converging on local minima, in 
practice it is sufficient to run expectation-maximization from multiple
initialisations in order to ensure that the global minima is reached.
At the maximization step we first update our estimate of the relative weights 
$\vec\weight\nextstep$ given the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
	% Include proof.
\end{equation}
\noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.

@@

The Fisher information for $\factorloads$ \todo{is hard to compute. I have 
the first derivative from the log likelihood function, but the second
derivative is (surprisingly) causing some trouble.}
Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
	&=& -\log{p(\factorloads)} + \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}

Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi) \nonumber \\
	&& +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData} \nonumber \\
	&& +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| - \frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| \nonumber \\
	&& +\textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents| \nonumber \\
	&& -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) + \frac{Q}{2}\log\kappa_q \nonumber \\
	&& +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} \nonumber \\
	&& \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} .
\end{eqnarray}

@@

In this experiment we use detailed chemical abundances of metal-poor stars 
\citep{Barklem:2005} to demonstrate the utility and limitations
in interpreting the latent factors as nucleosynthetic processes. The data 
can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), and those 14 chemical elements trace multiple nucleosynthetic pathways. Specifically the chemical
abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process). 
The data vector $\data$ includes [Fe/H] abundance ratios and all other elements
relative to Fe (e.g., [Ti/Fe]) after subtracting the mean value to allow the
latent factors to describe the abundance patterns away from the mean.

We performed a grid search for the optimal number of latent factors 
$\NumLatentFactors$ and components $\NumComponents$, starting from 
$\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
$\NumLatentFactors = 7$ and $\NumComponents = 3$. 
% Although it is common
%to add a regularization term to the diagonals of covariance matrices
%when performing expectation-maximization, we found that our existing
%grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
%and $\NumComponents$, and chose not to add any regularization term.
We initialised
the model parameters randomly for each permutation of $J$ and $K$, and continued
the expectation-maximization algorithm
until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood,
the BIC, and the message length recorded for the optimised set of model parameters 
for each permutation of $J$ and $K$. The negative log likelihood continues to
increase with increasing numbers of clusters, or increasing numbers of latent
factors.
The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
six latent factors are preferred, with no clustering. The shortest message length
is also found with no clustering,  but with four latent factors ($J = 4$, $K = 1$).
We adopt the message length as our objective function and chose $J = 4$ and $K = 1$
as the preferred model.

@@

%\todo{For example,
%the first factor does show positive contributions to Al, like the target load,
%but it also shows a slight negative contribution with respect to alpha elements
%(Mg, Ca)}, and a gradual increase among Fe-peak elements up to Ni. Similarly while
%Ca and Mg are indeed positive in the second factor, as targeted, there is a
%weak negative contribution to Mn abundances, which was not part of the target
%loads. 
Despite the caveats in interpreting latent factors as nucleosynthetic processes
(e.g. see Sections \ref{sec:exp-1} and \ref{sec:discussion}), here we will
qualitatively describe the latent factors found in our selected model.
The first latent factor $|\mathbf{L_0}|$ shows a non-zero contribution to most
elements (Figure~\ref{fig:exp2-factor-loads}). The elements Ca, Sc, Ti, Cr, and Co, mostly
Fe-peak elements, have the largest contribution from this factor. We interpret this latent
factor as contributions from core-collapse supernovae.
The second latent factor $|\mathbf{L_1}|$ has negligible contribution at most elements,
with strong contributions for C, and the neutron-capture elements Sr, Y, Ba, and Eu. We
interpret this latent factor as being most closely representative of the slow neutron
capture process. The third latent factor has almost no contribution to Ba, but has
non-zero contributions to other neutron-capture elements Sr, Y, and Eu. Since Eu has
the strongest contribution by $|\mathbf{L_2}|$, we interpret this factor as being most
representative of the rapid neutron capture process. The last latent factor $|\mathbf{L_3}|$
has a noticeable peak at Fe and the surrounding Fe-peak elements, with small non-zero contributions
at most elements.

@@

After accounting for the contributions by the latent factors and scores,
the specific scatter remaining in each dimension is shown in
Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
as 0.05~dex (Mg, Ca, Sc, Ti) up to 0.3~dex (C). It is perhaps not surprising that
the largest specific scatter is found in C, since the present day carbon abundance
is mostly affected by the evolutionary state of the star, rather than the natal
abundance. After C, the elements with large specific scatter include Al, Fe, Ni, 
and Ba. A large relative scatter in Al and Ba might suggest that there are other latent
factors that contribute to those elements, but those latent factors were not included because
the message length preferred a fewer number of factors. However, the scatter in 
Fe and Ni is somewhat surprising. Those elements usually have strong and numerous atomic
lines that can be easily measured in metal-poor stars.

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
	\caption{Specific scatter remaining in the \citet{Barklem:2005} data used in
			 Experiment~2 after accounting for the contributions by all latent
			 factors.}
    \label{fig:exp2-specific-scatter}
\end{figure}

@@

In this experiment we perform blind chemical tagging using the 
photospheric abundances released as part of the second \Galah\ 
data release \citep{Buder:2018}. This data set includes
up to 23 chemical abundances reported for 342,682 stars.
Some abundances are not reported, for a variety of 
reasons related to astrophysics or the adopted analyses
(e.g., a star is too hot to estimate the abundance for a given
transition, or the signal-to-noise of the spectrum is too low).
Our method as described cannot handle missing data entries for
a given star, and as such we are forced to exclude stars that
do not have a full complement of abundances. However, we also
chose to exclude some elemental abundances from our chemical
tagging experiments. For example, our experiments
with \Galah\ data do not include lithium or carbon abundances because the
photospheric varies throughout a star's lifetime (e.g., see Experiment 2). This is true
to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
but for the purposes of this experiment we assume that all other
photospheric abundances remain constant throughout  a star's
lifetime.

@@

We first selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances are untrustworthy. We then took all
stars with no erroneous flags in the following elemental abundance
measurements: Na, Mg, Al, Si, K, Ca, Sc, Ti, Cr, Mn, Fe, Co, Ni, Y, Ba, and Eu.
These elements are produced from multiple nucleosynthetic
pathways, and it is this subsample that we will use for 
our initial chemical experiment with \Galah\ data.
There were 1,153 stars that met these criteria.

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-specific-scatter.pdf}
	\caption{Specific scatter remaining in the \Galah\ data \citep{Buder:2018}
			 in Experiment~3 after accounting for the contributions by all
			 latent factors.}
    \label{fig:exp9-specific-scatter}
\end{figure}

@@

The relative weighting of the two components identified by this model using
\Galah\ data is nearly equal. In Figure~\ref{fig:exp9-latent-space} we show 
abundance ratios colored by their inferred component association.
It is clear that the component association in latent space does not directly
translate to a clear separation in data space: in this [Fe/H]-[Mg/Fe] plane
there is overlap in points between the two associations. 
\todo{When we look at latent space these components are more clearly separated.}
However, in general
the two components could be identified as high and low-$\alpha$ element ratio
components.

\subsection{Experiment~4:\\Including missing data from the \Galah\ survey} \label{sec:experiment-4}

\todo{Do we bother?}

@@

\todo{data-driven model of nucleosynthesis}

%We have presented a novel approach to chemical tagging that incorporates
%latent factors common to all stars, and allows for clustering in the relative
%contributions of those factors. This is analogous to simultaneously
%inferring nucleosynthetic yields as part of chemical tagging.

Earlier studies investigating the limitations of chemical tagging have shown
that the effective dimensionality of chemical abundance space is lower than
the number of reported abundances \citep[e.g.,][]{Price-Jones:2018}. 
This is expected. Theoretical models of stellar evolution and nucleosynthesis 
suggest that there are of order \todo{5} nucleosynthetic mechanisms that are
predominately responsible for the production of the $\sim$20 or so observed 
reported in those studies. As a consequence we can expect that the observed
chemical abundances are correlated in some way (e.g., by a linear combination
of a fewer number of nucleosynthetic pathways).  However, most approaches to
chemical tagging to date have assumed that the observed chemical abundances
are independent of each other \todo{\citep[e.g.,][]{who}}. Those inferences
will be biased by that assumption in a non-trivial way. This problem cannot be
resolved by simply lowering the number of abundances used for chemical tagging (e.g.,
down to something near the effective dimensionality), either, because those
chemical abundances are still produced by multiple pathways.

Here we have introduced a model to simultaneously account for the lower
effective dimensionality of chemical abundance space, and perform clustering
in that lower dimensional space. Although our model allows us to estimate
those latent factors that contribute to all stars, and cluster those stars by
their relative contributions from those factors, there are issues with the
model we adopt. The model we describe is very likely \emph{not} the correct
model to use to describe chemical abundances of stars. We require the latent
factors in our model to be mutually orthogonal to each other. This suggests
an astrophysical context where the mean nucleosynthetic yields (over all star
formation histories and stellar masses) of various
processes (e.g., $r$-process, $s$-process) are mutually orthogonal to each
other. Clearly this assumption is likely to be incorrect: the nuclear physics
of one environment where elements are produced  can be very different from
other environments, and there is no astrophysical constraint that those
latent factors should be mutually orthogonal.
A more flexible data-driven model of nucleosynthetic yields (as a function of mass,
stellar metallicity, and other factors) that does not require the latent
factors to be mutually orthogonal would be a worthy extension to this work.

There are other issues in our model that relate to our assumption of mutual
orthogonality. Even if nucleosynthetic yields were truely mutually orthogonal,
then the latent factors we infer are only \emph{identifiable} up until an
orthogonal basis. As we have seen in our experiments, the ordering and sign 
of the latent factors is not described \emph{a priori}. That means that we
must `assign' the latent factors we infer as being described by an astrophysical
process (e.g., the zeroth-latent factor is r-process). A more general limitation
of this is that the latent factors can be multiplied by some arbitrary rotation
matrix, leading to latent factor loads that are very different from what was
estimated by the model, but still lead to the exact same data (or log likelihood,
or Kullback-Liebler divergence, etc). As a consequence, we can only `identify'
latent factors up until this rotation. We have sought to address this by constructing
rotation matrices where the entries each latent factor correspond to our expectations
from astrophysical processes (whilst remaining orthogonal), but here we are limited
by what astrophysical processes we are \emph{expected} to find.

>>>>

\shorttitle{A data-driven model of nucleosynthesis with chemical tagging in latent space}
\shortauthors{Casey et al.}

@@

\title{A data-driven model of nucleosynthesis with chemical tagging in latent space}

@@

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated chemical abundances as being 
statistically independent, despite the nucleosynthesis literature showing that
most elements can be produced by more than one nucleosynthetic pathway.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is represented by a mixture model of multivariate normal distributions
in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the relative weighting (e.g., partial assignment) of each star
to each cluster, and to infer the latent factors that are common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We perform experiments on generated and real data, including literature chemical abundances
of metal-poor stars, and those from the second \Galah\ data release. 
We find that \todo{five} latent factors are preferred to explain the detailed 
abundances of metal-poor stars, and \todo{eight} are preferred given \todo{X} abundances
in \Galah\ data. In both experiments we recover latent factors that 
are representative of mean yields expected from the rapid- and slow-neutron
capture process, as well as nucleosynthesis in Type II supernovae. 
We find \todo{two} clusters are preferred in the \Galah\ data of 16 abundances
and 1,136 stars, which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
We expect more components
would be preferred given complete abundance inventory of a larger sample of stars.
We release accompanying software that scales well with the available data,
allowing model parameters to be optimised in a few minutes even given $\sim$$10^{7.5}$ 
abundance measurements.
%We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
%enabling model parameter estimation in minutes given $\sim$$10^7$
%chemical abundance measurements.
\end{abstract}

@@

Here we present a new approach to chemical tagging that allows us to identify 
the latent (unobserved) factors that contribute to the chemical abundances of 
all stars (e.g., nucleosynthetic yields) while simultaneously performing 
clustering in the latent space. Notwithstanding caveats that we will
discuss in detail, this allows us to infer nucleosynthetic yields rather than
strictly prescribe them from models. Moreover, the scale of the clustering
problem reduces by a significant fraction because the clustering is performed in
a lower dimensional latent space instead of the higher dimensional data space.
In Section~\ref{sec:methods} we describe the model and the methods we use to
estimate the model parameters. Section~\ref{sec:experiments} describe the 
experiments performed using generated and real data sets. We discuss the results
of these experiments in Section~\ref{sec:discussion}, including the caveats with
the model as described. We conclude in Section~\ref{sec:conclusions}.

@@

At the maximization step we update our estimates of the parameters,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
The updated parameters estimates are found by setting the second derivative
of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
the parameter values\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our chosen objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this difference is inconsistency becomes serious at small $N$.}
In doing so this guarantees that every updated
estimate of the model parameters is guaranteed to increase the log likelihood.
Although there are no guarantees against converging on local minima, in 
practice it is sufficient to run expectation-maximization from multiple
initialisations in order to ensure that the global minima is reached.
At the maximization step we first update our estimate of the relative weights 
$\vec\weight\nextstep$ given the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
	% Include proof.
\end{equation}
\noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.

@@

The Fisher information for $\factorloads$ \todo{is hard to compute. I have 
the first derivative from the log likelihood function, but the second
derivative is (surprisingly) causing some trouble.}
Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
	I(\factorloads) &=& -\log\left(\frac{p(\factorloads)}{\sqrt{|\mathcal{F}(\factorloads)|}}\right) \nonumber \\
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} \nonumber \\
	&& \cdots \,\, + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
	&& \cdots \todo{+\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:prior_L}
\end{eqnarray}

Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
\begin{widetext}
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
 +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
 +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2} + \todo{\frac{1}{2}\log{|\mathcal{F}(\factorloads)|}} \quad . \label{eq:message-length}
 \end{eqnarray}
\end{widetext}

@@

In this experiment we use detailed chemical abundances of metal-poor stars 
from \citet{Barklem:2005} to illustrate the utility and limitations
in interpreting the latent factors as nucleosynthetic processes. The data 
can be described as 14 chemical abundance measurements from 61 stars ($\NumData = 61$, $\NumDimensions = 14$), and those 14 chemical elements trace multiple nucleosynthetic pathways. Specifically the chemical
abundances measured include Al (light odd-Z), Mg and Ca ($\alpha$-process), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak elements), as well as Sr, Y, Ba, and Eu ($s$/$r$-process). 
The data vector includes [Fe/H] abundance ratios and all other elements
relative to Fe (e.g., [Ti/Fe]) after subtracting the mean value in each dimension.

We performed a grid search for the optimal number of latent factors 
$\NumLatentFactors$ and components $\NumComponents$, starting from 
$\NumLatentFactors = 1$ and $K = 1$, and trialling up until 
$\NumLatentFactors = 7$ and $\NumComponents = 3$. 
% Although it is common
%to add a regularization term to the diagonals of covariance matrices
%when performing expectation-maximization, we found that our existing
%grid search boundaries were sufficient to find the preferred $\NumLatentFactors$
%and $\NumComponents$, and chose not to add any regularization term.
We initialised
the model parameters randomly for each permutation of $J$ and $K$, and continued
the expectation-maximization algorithm
until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood,
the BIC, and the message length recorded for the optimised set of model parameters 
for each permutation of $J$ and $K$. The negative log likelihood continues to
increase with increasing numbers of clusters, or increasing numbers of latent
factors.
The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
five latent factors are preferred with no clustering. The shortest message length
is also found with no clustering,  but with four latent factors ($J = 4$, $K = 1$).
We adopt the message length as our objective function and chose $J = 4$ and $K = 1$
as the preferred model.

@@

%\todo{For example,
%the first factor does show positive contributions to Al, like the target load,
%but it also shows a slight negative contribution with respect to alpha elements
%(Mg, Ca)}, and a gradual increase among Fe-peak elements up to Ni. Similarly while
%Ca and Mg are indeed positive in the second factor, as targeted, there is a
%weak negative contribution to Mn abundances, which was not part of the target
%loads. 
Despite the caveats in interpreting latent factors as nucleosynthetic processes
(e.g., see Sections~\ref{sec:exp-1} and \ref{sec:discussion}), here we will
qualitatively describe the latent factors found in our selected model.
The first latent factor $|\mathbf{L_0}|$ shows a non-zero contribution to most
elements (Figure~\ref{fig:exp2-factor-loads}). The elements Ca, Sc, Ti, Cr, and Co, mostly
Fe-peak elements, have the largest contribution from this factor. We interpret this latent
factor as being most similar to expected contributions from core-collapse supernovae.
The second latent factor $|\mathbf{L_1}|$ has negligible contribution at most elements,
with strong contributions for C, and the neutron-capture elements Sr, Y, Ba, and Eu. We
interpret this latent factor as being most closely representative of the slow neutron
capture process, which primarily occurs in asymptotic giant branch stars. The third latent factor has almost no contribution to Ba, but has
non-zero contributions to other neutron-capture elements Sr, Y, and Eu. Since Eu has
the strongest contribution by $|\mathbf{L_2}|$, we interpret this factor as being most
representative of the rapid neutron capture process. The last latent factor $|\mathbf{L_3}|$
has a noticeable peak at Fe and the surrounding Fe-peak elements, with small non-zero contributions at most elements. \todo{What to say of $|\mathbf{L_3}|$?}

@@

After accounting for the contributions by the latent factors and scores,
the specific scatter remaining in each dimension is shown in
Figure~\ref{fig:exp2-specific-scatter}. The remaining scatter varies from as low 
as 0.05~dex (Mg, Ca, Sc, Ti) up to 0.3~dex (C). It is perhaps not surprising that
the largest specific scatter is found in C, since the present day carbon abundance
is mostly affected by the evolutionary state of the star, rather than the natal
abundance. After C, the elements with large specific scatter include Al, Fe, Ni, 
and Ba. A large relative scatter in Al and Ba might suggest that there are other latent
factors that contribute to those elements, but those latent factors were not included because
the message length preferred a fewer number of factors, or it could be limitations in
the assumptions of our model (e.g., we assume mutually orthogonal latent factors, an assumption that is likely provably false; see Section~\ref{sec:discussion}).
%However, the scatter in 
%Fe and Ni is somewhat surprising. Those elements usually have strong and numerous atomic
%lines that can be easily measured in metal-poor stars.

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-log-xfe-offset-specific-scatter.pdf}
	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \citet{Barklem:2005} data used in
			 Experiment~2 after accounting for the contributions by all latent
			 factors.}
    \label{fig:exp2-specific-scatter}
\end{figure}

@@

In this experiment we perform blind chemical tagging using the 
photospheric abundances released as part of the second \Galah\ 
data release \citep{Buder:2018}. This data set includes
up to 23 chemical abundances reported for 342,682 stars.
Some abundances are not reported for a variety of 
reasons related to astrophysics or the adopted analyses
(e.g., a star is too hot to estimate the abundance for a given
electronic transition, or the signal-to-noise of the spectrum is too low).
Our method as described cannot efficiently handle missing data 
entries for a given star, and as such we are forced to exclude stars that
do not have a full complement of abundances. However, we also
chose to exclude some elemental abundances from our chemical
tagging experiments. For example, our experiments
with \Galah\ data do not include lithium or carbon abundances because the
photospheric varies throughout a star's lifetime (e.g., see Experiment 2). This is true
to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
but for the purposes of this experiment we assume that all other
photospheric abundances remain constant throughout a star's
lifetime.

@@

We first selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances would be untrustworthy. We then took all
stars with no erroneous flags in the following elemental abundance
measurements: Na, Mg, Al, Si, K, Ca, Sc, Ti, Cr, Mn, Fe, Co, Ni, Y, Ba, and Eu.
These elements are produced from multiple nucleosynthetic
pathways, and it is this subsample that we will use for 
our initial chemical experiment with \Galah\ data.
There were 1,153 stars that met these criteria.

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp9-log-xfe-offset-specific-scatter.pdf}
	\caption{Specific scatter $\sqrt{\specificvariance}$ remaining in the \Galah\ data \citep{Buder:2018}
			 in Experiment~3 after accounting for the contributions by all
			 latent factors.}
    \label{fig:exp9-specific-scatter}
\end{figure}

@@

The two components identified by this model have nearly equal mixing proportions.
In Figure~\ref{fig:exp9-latent-space} we show the latent space coloured by the
inferred component association, and in Figure~\ref{fig:exp9-cluster} we project
this to real space showing [Fe/H] abundance ratios relative to [Mg/Fe].
It is clear that the component association in latent space does not directly
translate to a clear separation in data space: in this [Fe/H]-[Mg/Fe] plane
there is overlap in points between the two associations. 
However, when we consider the clustering in latent space these components 
are more clearly separated. In general the two components identified by the
model could be described as high- and low-$\alpha$ element ratio stars in
the disk of the Milky Way.

With a larger sample of stars that have a complete abundance inventory we would expect to 
identify more components, and perhaps unique star formation events. However
because our model requires no missing abundances in any element, we would
need to reduce the number of elemental abundances we include in order to increase
the sample of stars. And by doing so we restrict the number of latent factors
that can be inferred from the data.

@@

Literature studies testing the limitations chemical tagging have shown
that the effective dimensionality of chemical abundance space is lower than
the number of reported abundances \citep[e.g.,][]{Price-Jones:2018}. 
This is expected. Theoretical models of stellar evolution and nucleosynthesis
would suggest that there are of order at most ten important nucleosynthetic 
contributors that are responsible for the production of the $\sim$20--30
chemical abundances typically reported. As a consequence we can expect that 
the observed chemical abundances are correlated in some way (e.g., by some
linear combination of a fewer number of nucleosynthetic pathways).  However, 
most approaches to chemical tagging to date have assumed that the observed 
chemical abundances are independent of each other. This problem cannot be
resolved by simply lowering the number of abundances used for chemical tagging (e.g.,
down to something near the effective dimensionality), either, because those
chemical abundances are still produced by multiple pathways.

Here we have introduced a model to simultaneously account for the lower
effective dimensionality of chemical abundance space, and perform clustering
in that lower dimensional space. This provides a data-driven model of
nucleosynthesis yields and chemical tagging that allows us to simultaneously
estimate the latent factors that contribute to all stars, and cluster those 
stars by their relative contributions from each factors. The results are
encouraging in that we find latent factors that are representative of the
expected yields from dominant nucleosynthetic channels. However, the model that
we describe is very likely \emph{not} the correct model to use to represent 
chemical abundances of stars. Here we discuss the limitations of our model 
in detail.

We require latent factors to be mutually orthogonal to in order to resolve
the indeterminacy in our model. This suggests an astrophysical context where 
the mean nucleosynthetic yields (integrated over all stellar masses and star
formation histories) of various nucleosynthetic processes (e.g., $r$-process, 
$s$-process) are mutually orthogonal to each other. Clearly this assumption 
is likely to be incorrect: the nuclear physics
of one environment where elements are produced will be very different from
others, and there is no astrophysical constraint that those
yields (or latent factors) should be mutually orthogonal.
One could represent them by a hierarchical data-driven model where the yields contribute as a function of stellar mass, 
metallicity, and other factors), but in principle to resolve the indeterminacy
the mean of these yields would require mutual orthogonality. Introducing a constraint on the factor scores that resolves this indeterminacy and allows for more flexible latent factors would be a worthy extension to this work.

The constraint of mutual orthogonality limits the inferences we want to make
about stellar nucleosynthetic yields. For example, after accounting for all
known sources of potassium production in the Milky Way, galactic chemical evolution
models under-predict the level of potassium in the Milky Way by more than an
order of magnitude \citep{Kobayashi:2006}. From our inferences in Experiment~3
using \Galah\ data, we find multiple latent factors have non-zero contributions
to potassium: $\factorloads_1$, $\factorloads_2$, and $\factorloads_3$. Let us consider
$\factorloads_3$, which is the dominant contributor of the principally $r$-process element europium. If our model confidently and
reliably associated the production of potassium with the rapid neutron capture
process, it could help explain the peculiar abundances of stars enhanced in
potassium and depleted in magnesium \citep{Mucciarelli:2012,Cohen:2012} -- a chemical
abundance pattern that currently lacks explanation \citep{Kemp:2018}.
However, is the potassium contribution that we infer physically realistic,
or is it a consequence of requiring that the latent factors are mutually
orthogonal? Distinguishing these possibilities is non-trivial, which is in
part why caution is warranted when trying to interpret latent factor models.
In this situation it is worth commenting that potassium has the largest specific
scatter (Figure~\ref{fig:exp9-specific-scatter}), suggesting that the contributions
of potassium are not as well described by the latent factor model as other
elements.
